Phenotype and risk factors of venom-induced anaphylaxis: a case - control study of the European Anaphylaxis Registry

### 2019-12-19 analysis/paper/paper.Rmd
gramarly corrections

---
title: "Phenotype and risk factors of venom induced anaphylaxis: a case - control study of the European Anaphylaxis Registry"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    #bookdown::html_document2:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/american-medical-association.csl" # Insert path for the bib-style
---

@@

- VIA - venom induced anaphylaxis
- BST - baseline Serum Tryptase
- EAI - epinephrine autoinjector
- MCAS - mast cell activation syndrome
- ER - emergency room

@@

**Findings:** This case-control study showed that venom induced anaphylaxis significantly more often presented with cardiovascular symptoms. Secondly, severe cases of venom induced anaphylaxis were significantly more often associated with the lack of skin symptoms upon anaphylaxis and showed a significant interaction with the levels of baseline serum tryptase (in range from 8 - 11 ng/ml).

@@

**Importance:** Venom induced anaphylaxis is a common, potentially life-threatening hypersensitivity reaction associated with specific: 1) symptom profile, 2) cofactors, and 3) management. Identifying the differences in phenotypes of anaphylaxis is crucial for future management guidelines and the development of a personalized medicine approach.

**Objective:** This study aimed to evaluate the phenotype and risk factors of venom induced anaphylaxis in comparison to other elicitors of anaphylaxis.

**Design:** Using data from the European Anaphylaxis Registry (`r nrow(data)` cases) we identified `r length(which(data4$d_elicitor_gr5== "insects" & data4$reaction_type_brown=="anaphylaxis"))` with venom induced anaphylaxis and analyzed these in comparison to sex-and-age-matched anaphylaxis cases triggered by other elicitors (non-VIA n = `r nrow(age_sex_matched %>% filter(grouping == "other"))`).

@@

**Participants:** All patients with the diagnosis of moderate to severe anaphylaxis (according to Ring and Messmer grade II-IV) who gave an informed consent to provide their clinical data for the Anaphylaxis Registry were included in the study. Subsequently, we matched the group of patients presenting with venom anaphylaxis to a sex and age matched group of patients who presented with anaphylaxis to other elicitors (i.e. food, drugs).

**Main Outcome(s) and Measure(s):** This study measured severity of anaphylaxis on the Ring and Messmer scale. Hypotheses regarding the comparisons between groups were formulated during and after data collection.

**Results:** Venom induced anaphylaxis more frequently involved more than three organ systems and was associated with cardiovascular symptoms. The absence of skin symptoms during anaphylaxis correlated with baseline serum tryptase and was associated with an increased risk of a severe reaction. Intramuscular or intravenous epinephrine was administered significantly less often in venom induced anaphylaxis, in particular in patients without prior history of anaphylaxis. Baseline serum tryptase within upper normal range (8-11.5 ng-ml) was more frequently associated with severe anaphylaxis.

**Conclusions and Relevance:** Regarding the fact that venom induced anaphylaxis frequently affects the cardiovascular system, patients were undertreated with epinephrine. The lack of skin symptoms (i.e. urticaria or flushing) during anaphylaxis and baseline serum tryptase levels within upper normal limits were associated with severe reactions. Patients with serum tryptase of above 8 ng/ml and a history of venom anaphylaxis may require additional prophylaxis.

@@

Hypersensitivity to insect venom presents as a systemic reaction (anaphylaxis) in up to 0.3--7.5% of the adult population [@Bil2008]. Venom induced anaphylaxis (VIA) can be fatal, and patients sometimes require lifelong specific immunotherapy [@Sturm2017]. There is a need for a more precise identification of biomarkers, and better definition of phenotypes of anaphylaxis [@JimenezRodriguez2018]. Also, in order to facilitate a precision-medicine approach [@Muraro2017] for the diagnosis of anaphylaxis, a better understanding of its clinical phenotypes is required.

Anaphylaxis is a clinical diagnosis with a variety of triggering factors and clinical presentations. Symptom profiles and specific cofactors for venom-induced anaphylaxis (VIA) had previously been analyzed in an uncontrolled manner albeit in relatively small cohorts [@LANTNER; @Ruff2009; @nittner2015insect].

@@

The diagnosis of anaphylaxis was based on the definition by NIAID/FAAN [@Sampson2006] and the severity according to the Ring and Messmer Scale [@Ring1977]. Grades III and IV (presenting with significant hypoxia, hypotension, confusion, and loss of consciousness, or incontinence or cardiac arrest) were considered severe. Mastocytosis patients were defined as having a documented diagnosis of mastocytosis prior to the reaction. The Registry is designed for reporting cases of moderate to severe anaphylaxis (Ring and Messmer grades II - IV).

Due to the large number of documented reactions in the European Anaphylaxis Registry, we were able to match the VIA with non-VIA cases according to sex and age to reduce the comparison bias. 
When we analyzed a density plot of VIA cases according to age, we determined a bimodal distribution forming two subsets of patients with a cutoff age of 22 (Fig. \@ref(fig:flow)B). Subsequently, we compared the management in both groups and matched the control group according to the severity of a reaction. Propensity score matching has been performed using the "MatchIt" package for R [@matchit]. The results of the propensity score matching are illustrated in Fig. \@ref(fig:flow)B-D and supplementary Fig \@ref(fig:matchingresult).

The final database included `r countries %>% summarise(sum(n))` cases of venom induced anaphylaxis reported from allergy centers in `r countries %>% summarize(length(n))` countries and sex-and-age matched control group. 
We compared the frequency of various symptoms, cofactors --- known to increase the risk of severe anaphylaxis, [@Worm2018], and management in both groups.

Based on the severity and symptom profile, and previous reports [@Ruff2009] we defined sub-elevated baseline serum tryptase (BST) values as 8 - 11.5 ng/ml (Fig. \@ref(fig:figskinsymptoms)B).

<!-- ## Data handling and statitics -->
We used the R Statistical Package [@R] for statistical analysis. A simple comparison of categorical variables was performed using either Chi^2^ test or Fisher's exact test (where the number of observations in a bin was less than 10). Continuous variables were analyzed using Mann-Whitney U test. In case of comparisons with two or more independent variables, we used Factorial ANOVA or Generalized Linear Models. We defined statistical significance as α = 0.05. Data, along with the analysis script, can be accessed online at https://github.com/wolass/venomanaphylaxiscompendium.

We developed a random forest classifier (using the "randomForest" package for R [@rf]) in order to find therapeutic approaches that varied the most between VIA / non-VIA group and presented the results as Gini importance [@Strobl2007]. Moreover, an association analysis of therapeutic interventions and symptoms was performed. The resulting phi values were scaled and presented in a heatmap with automatic clustering using Ward’s Agglomerative Hierarchical Clustering with Euclidean distances [@heatmaply].

@@

We found that `r skin_symptoms$masto_tab[2,3]` (`r (skin_symptoms$masto_tab$prop[c(2)]*100) %>% round(1)`%) of patients with concomitant mastocytosis had anaphylaxis without skin symptoms (i.e. urticaria and flushing) which was significantly more frequent compared to patients without diagnosed mastocytosis (`r skin_symptoms$masto_tab[1,3]`; `r (skin_symptoms$masto_tab$prop[c(1)]*100) %>% round(1)`%, `r pval(skin_symptoms$masto_chisq$p.value)`). This finding was most prominently seen in VIA (Fig. \@ref(fig:figskinsymptoms)A).

Similarly, in non-mastocytosis patients undergoing VIA, skin symptoms (i.e. urticaria or flushing) were less often present than if anaphylaxis was triggered by other elicitors (`r skin_symptoms$tab[2,3]`; `r (skin_symptoms$tab$prop[2]*100) %>% round(1)`% vs `r skin_symptoms$tab[4,3]`; `r (skin_symptoms$tab$prop[4]*100) %>% round(1)`% respectively, `r pval(skin_symptoms$chisq$p.value)`). Moreover, in this specific subgroup of patients (i.e. non-mastocytosis patients lacking skin symptoms) VIA was significantly more frequently severe (`r skin_symptoms$severe_tab[c(6),4]`,`r (skin_symptoms$severe_tab$prop[c(6,2)]*100) %>% round(1) %>% paste0(collapse="% in VIA vs. ")`%, `r skin_symptoms$severe_tab[c(2),4]`; `r pval(skin_symptoms$masto_chisq$p.value)`,  Fig. \@ref(fig:figskinsymptoms)B).

@@

Concomitant cardiovascular diseases were more prevalent in VIA than in non-VIA cases (`r age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% mutate(prop= n/sum(n)) %>% {.[c(2,5),4]} %>% pull() %>% compare(x = (age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% {.[c(2,5),3]} %>% pull()), type = "both")`) and were associated with a higher risk of severe anaphylaxis when elicited by insects but were not relevant in non-VIA cases (Fig. \@ref(fig:figcofactors)). Interestingly, BST values were increased in patients with concomitant cardiovascular diseases, irrespectively of the reaction severity (Fig. S\@ref(fig:trypcardiocur)).

@@

The effect of using ACE-I (as well as beta-blockers) on the risk of severe anaphylaxis correlated with the coexisting cardiovascular diseases. ACE-I use was, however, more often associated with cardiac arrests in all anaphylaxis cases (`r ace_tab %>% filter(q_114_cardiac_arrest =="yes") %>% dplyr::select(prop) %>% pull() %>% rev() %>% compare( x = ace_tab %>% filter(q_114_cardiac_arrest =="yes") %>% dplyr::select(n) %>% pull() %>% rev(), type = "both")`, `r pval(ace_tab %>% dplyr::select(-prop) %>% spread(q_114_cardiac_arrest,n) %>% {chisq.test(.[,2:3])$p.value})`) regardless of the elicitor  (Fig. \@ref(fig:figcofactors)C). 
Beta-blocker use was associated with a higher severity of anaphylaxis and with the onset of cardiovascular symptoms (cardiac arrest, chest pain), but was comparable between both VIA and non-VIA`r summary(beta_fit)[["coefficients"]][3,4] %>% pval()`). Surprisingly, arrhythmia was more frequently reported in patients with VIA and concomitant beta-blockers (Fig. \@ref(fig:figcofactors)C).

@@

`r table(rdb$grouping,rdb$q_160_ever_react)[,1:2][1,2]` (`r treact[1,2] %>% round(1)`%) of patients with insect allergy had experienced venom anaphylaxis in the past. If the reaction was elicited by other elicitors (i.e. non-VIA) --- previous reactions were more frequently seen `r table(rdb$grouping,rdb$q_160_ever_react)[,1:2][2,2]` (`r treact[2,2]%>% round(1)`%, `r p %>% pval()`).  We observed `r sum(repeated_reaction_in_Reg)` patients with at least two fully-documented reactions. Out of these `r sum(greater_same[,2])` (`r (sum(greater_same[,2])/sum(repeated_reaction_in_Reg)*100)%>% round(1)`%) had insect elicited anaphylaxis and in `r greater_same[1,2]` of them (`r (greater_same[1,2]/sum(greater_same[,2])*100 )%>%round(1) `%) the following reaction was more severe than before. In `r greater_same[3,2]` (`r (greater_same[3,2]/sum(greater_same[,2])*100) %>% round(1)`%) cases the reaction was similar in severity.

@@

We evaluated epinephrine use (administered by any route from patients themselves and medical professionals) in both ambulatory and emergency room settings.   
Patients who underwent VIA significantly less often received epinephrine treatment than in other anaphylaxis cases (26.9% vs 34.6%, `r test_adrenuse1$p.value %>% pval()`). After adjusting both groups for similar severity - the difference in epinephrine use was still significant irrespective of the administration route (`r ttadren1`, Fig \@ref(fig:adrenalineuse)B). 
<!-- Antihistaminic drugs, on the other hand, were given more often in cases of VIA when compared to non-VIA. -->
```{r }
pvaladrenprev <- adren_prev$data %>% 
  spread(key = d_520_adren1,value = n) %>% 
  filter(q_160_ever_react=="reacted previously") %>% 
  {chisq.test(.[,3:4])$p.value} %>% pval()

@@

A positive history of anaphylaxis influenced the therapy of a current episode as well. Epinephrine as a first-line treatment was given less often in VIA cases when compared to other cases **if patients did not report a previous history of anaphylaxis** (`r pvaladrenprev_no`), but in patients reporting previous reactions, there was no difference in epinephrine therapy (`r pvaladrenprev`, Fig. \@ref(fig:adrenalineuse)B). 
Similarly, there were no differences in the epinephrine use between VIA and non-VIA when only severe reactions were taken into consideration (`r adren_severity_sev`). However, when we restricted the analysis to moderate anaphylaxis cases --- non-VIA patients received epinephrine more frequently than VIA (`r adren_severity_mild`).  Presence of skin symptoms during these mild reactions also was associated with lower fraction of epinephrine treated patients (Fig. \@ref(fig:rfplot)).

@@

Cofactors acn promote the onset and increase severity of anaphylaxis. 
Therefore we evaluated a variety of known cofactors regarding their impact to increase the risk of severe VIA. As expected, t

@@

The rate of concomitant cardiovascular diseases was higher in VIA than non-VIA. They are an essential cofactor increasing the risk of a severe reaction **if hymenoptera elicited the anaphylaxis**. This association was not significant in anaphylaxis elicited by other elicitors.  Notably, cardiac arrest occurred more frequently in patients with elevated BST (> 8 ng/ml), especially in VIA. 
Nevertheless, the pathomechanism promoting cardiovascular symptoms in VIA requires further investigation.

@@

Importantly, the absence of skin symptoms was associated with more severe VIA, which was still present after excluding patients with a known diagnosis of mastocytosis. Subsequently, the correlation of BST levels with severity of anaphylaxis lead us to identify an interaction between the absence of skin symptoms and VIA using generalized linear regression.

@@

Although there are no absolute contraindications for using epinephrine in anaphylaxis, one potential scenario where clinicians tend to be reluctant to using epinephrine is a hypersensitivity reaction presenting with high blood pressure and tachycardia, which may be present at the initial phase of VIA, due to a psychologic reflex. In theory, these less severe cases of VIA may display some form of stress-related blood pressure increase. As we lack data to confirm or discard this theory future analysis of this question is of great clinical value.

@@

Nevertheless, because the European Anaphylaxis Registry has until now gathered over 12,000 cases of anaphylaxis - it was possible to perform a case-controlled analysis on a relatively large number of cases and investigate many aspects of VIA. It is worth underlining the important function of international registries especially in diseases where targeted studies are not possible.

@@

# Acknowledgements

@@

```{r figsymptoms, fig.cap="Symptoms of venom induced anaphylaxis (VIA) compared to other elicitors. A: Proportional presentation of specific reaction symptoms in VIA and non-VIA according to cardiovascular (cardio.), gastroenterologic (gastro.), and respiratory (resp.) organ systems. B: High-level overview of involved organ systems and selected cofactors in the form of a radar plot. C: difference in symptoms of VIA among patients under 22 and over 22 years of age. * denotes significant differences between groups.", fig.width=8.5, fig.height = 6.2}
fig_symptoms_3
```

@@

```{r supplement, fig.width= 7, fig.height= 5, fig.cap="Symptoms of anaphylaxis. A: The association between cardiac arrest and concomitant mastocytosis in VIA and non-VIA. B:  Hypotension frequency in two age groups of anaphylaxis. C: Crammer's V as the measure of association between groups anaphylaxis (VIA vs. non-VIA). Higher values indicate stronger assotiation with IVA."}

@@

```{r rfplot, fig.cap="Therapy of anaphylaxis. A: Patients who presented with skin symptoms and VIA less often received epinephrine than if skin symptoms were absent during the reaction. B: Variable importance in the unsupervised classification between VIA and non-VIA using random Forest classifier.", fig.height=3.5, fig.width=7.1}
ggarrange(
  # adren_prev+
  #   geom_segment(aes(x = 0.5,
  #                    xend = 2.5,
  #                    y = 0.33,
  #                    yend = 0.33),
  #                color="white",
  #                linetype =2),
  skin_sympt_adrenaline,
  rf_ggplot_treatment,
  nrow = 1,
  ncol = 2,
  labels=c("A","B"),
  widths = c(1,1.3)
)
```

>>>>

---
title: "Phenotype and risk factors of venom-induced anaphylaxis: a case - control study of the European Anaphylaxis Registry"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    #bookdown::html_document2:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/american-medical-association.csl" # Insert path for the bib-style
---

@@

- VIA - venom-induced anaphylaxis
- BST - baseline Serum Tryptase
- EAI - epinephrine autoinjector
- MCAS - mast cell activation syndrome
- ER - emergency room

@@

**Findings:** This case-control study showed that venom-induced anaphylaxis significantly more often presented with cardiovascular symptoms. Secondly, severe cases of venom-induced anaphylaxis were significantly more often associated with the lack of skin symptoms upon anaphylaxis and showed a significant interaction with the levels of baseline serum tryptase (in range from 8 - 11 ng/ml).

@@

**Importance:** Venom-induced anaphylaxis is a common, potentially life-threatening hypersensitivity reaction associated with specific: 1) symptom profile, 2) cofactors, and 3) management. Identifying the differences in phenotypes of anaphylaxis is crucial for future management guidelines and the development of a personalized medicine approach.

**Objective:** This study aimed to evaluate the phenotype and risk factors of venom-induced anaphylaxis in comparison to other elicitors of anaphylaxis.

**Design:** Using data from the European Anaphylaxis Registry (`r nrow(data)` cases) we identified `r length(which(data4$d_elicitor_gr5== "insects" & data4$reaction_type_brown=="anaphylaxis"))` with venom-induced anaphylaxis and analyzed these in comparison to sex- and age-matched anaphylaxis cases triggered by other elicitors (non-VIA n = `r nrow(age_sex_matched %>% filter(grouping == "other"))`).

@@

**Participants:** All patients with the diagnosis of moderate to severe anaphylaxis (according to Ring and Messmer grade II-IV) who gave informed consent to provide their clinical data for the Anaphylaxis Registry were included in the study. Subsequently, we matched the group of patients presenting with venom anaphylaxis to sex- and age-matched group of patients who presented with anaphylaxis to other elicitors (i.e., food, drugs).

**Main Outcome(s) and Measure(s):** This study measured the severity of anaphylaxis on the Ring and Messmer scale. Hypotheses regarding the comparisons between groups were formulated during and after data collection.

**Results:** Venom-induced anaphylaxis more frequently involved more than three organ systems and was associated with cardiovascular symptoms. The absence of skin symptoms during anaphylaxis correlated with baseline serum tryptase and was associated with an increased risk of a severe reaction. Intramuscular or intravenous epinephrine was administered significantly less often in venom-induced anaphylaxis, in particular in patients without prior history of anaphylaxis. Baseline serum tryptase within the upper normal range (8-11.5 ng-ml) was more frequently associated with severe anaphylaxis.

**Conclusions and Relevance:** Regarding the fact that venom-induced anaphylaxis frequently affects the cardiovascular system, patients were undertreated with epinephrine. The lack of skin symptoms (i.e., urticaria or flushing) during anaphylaxis and baseline serum tryptase levels within the upper normal limits were associated with severe reactions. Patients with serum tryptase of above 8 ng/ml and a history of venom anaphylaxis may require additional prophylaxis.

@@

Hypersensitivity to insect venom presents as a systemic reaction (anaphylaxis) in up to 0.3--7.5% of the adult population [@Bil2008]. Venom-induced anaphylaxis (VIA) can be fatal, and patients sometimes require lifelong specific immunotherapy [@Sturm2017]. There is a need for more precise identification of biomarkers, and better definition of phenotypes of anaphylaxis [@JimenezRodriguez2018]. Also, in order to facilitate a precision-medicine approach [@Muraro2017] for the diagnosis of anaphylaxis, a better understanding of its clinical phenotypes is required.

Anaphylaxis is a clinical diagnosis with a variety of triggering factors and clinical presentations. Symptom profiles and specific cofactors for venom-induced anaphylaxis (VIA) had previously been analyzed in an uncontrolled manner, albeit in relatively small cohorts [@LANTNER; @Ruff2009; @nittner2015insect].

@@

The diagnosis of anaphylaxis was based on the definition by NIAID/FAAN [@Sampson2006] and the severity according to the Ring and Messmer Scale [@Ring1977]. Grades III and IV (presenting with significant hypoxia, hypotension, confusion, and loss of consciousness, or incontinence or cardiac arrest) were considered severe. Mastocytosis patients were defined as having a documented diagnosis of mastocytosis prior to the reaction. The Registry is designed for reporting cases of moderate to severe anaphylaxis (Ring and Messmer grades II-IV).

Due to a large number of documented reactions in the European Anaphylaxis Registry, we were able to match the VIA with non-VIA cases according to sex and age to reduce the comparison bias. 
When we analyzed a density plot of VIA cases according to age, we determined a bimodal distribution forming two subsets of patients with a cutoff age of 22 (Fig. \@ref(fig:flow)B). Subsequently, we compared the management in both groups and matched the control group according to the severity of a reaction. Propensity score matching has been performed using the "MatchIt" package for R [@matchit]. The results of the propensity score matching are illustrated in Fig. \@ref(fig:flow)B-D and supplementary Fig \@ref(fig:matchingresult).

The final database included `r countries %>% summarise(sum(n))` cases of venom-induced anaphylaxis reported from allergy centers in `r countries %>% summarize(length(n))` countries and sex- and age-matched control group. 
We compared the frequency of various symptoms, cofactors --- known to increase the risk of severe anaphylaxis, [@Worm2018], and management in both groups.

Based on the severity and symptom profile and the previous reports [@Ruff2009], we defined sub-elevated baseline serum tryptase (BST) values as 8 - 11.5 ng/ml (Fig. \@ref(fig:figskinsymptoms)B).

<!-- ## Data handling and statitics -->
We used the R Statistical Package [@R] for statistical analysis. A simple comparison of categorical variables was performed using either the Chi^2^ test or Fisher's exact test (where the number of observations in a bin was less than 10). Continuous variables were analyzed using the Mann-Whitney U test. In case of comparisons with two or more independent variables, we used Factorial ANOVA or Generalized Linear Models. We defined statistical significance as α = 0.05. Data, along with the analysis script, can be accessed online at https://github.com/wolass/venomanaphylaxiscompendium.

We developed a Random Forest classifier (using the "randomForest" package for R [@rf]) in order to find therapeutic approaches that varied the most between VIA / non-VIA group and presented the results as Gini importance [@Strobl2007]. Moreover, association analysis of therapeutic interventions and symptoms was performed. The resulting phi values were scaled and presented in a heatmap with automatic clustering using Ward’s Agglomerative Hierarchical Clustering with Euclidean distances [@heatmaply].

@@

We found that `r skin_symptoms$masto_tab[2,3]` (`r (skin_symptoms$masto_tab$prop[c(2)]*100) %>% round(1)`%) of patients with concomitant mastocytosis had anaphylaxis without skin symptoms (i.e., urticaria and flushing), which was significantly more frequent compared to patients without diagnosed mastocytosis (`r skin_symptoms$masto_tab[1,3]`; `r (skin_symptoms$masto_tab$prop[c(1)]*100) %>% round(1)`%, `r pval(skin_symptoms$masto_chisq$p.value)`). This finding was most prominently seen in VIA (Fig. \@ref(fig:figskinsymptoms)A).

Similarly, in non-mastocytosis patients undergoing VIA, skin symptoms (i.e., urticaria or flushing) were less often present than if anaphylaxis was triggered by other elicitors (`r skin_symptoms$tab[2,3]`; `r (skin_symptoms$tab$prop[2]*100) %>% round(1)`% vs. `r skin_symptoms$tab[4,3]`; `r (skin_symptoms$tab$prop[4]*100) %>% round(1)`% respectively, `r pval(skin_symptoms$chisq$p.value)`). Moreover, in this specific subgroup of patients (i.e., non-mastocytosis patients lacking skin symptoms) VIA was significantly more frequently severe (`r skin_symptoms$severe_tab[c(6),4]`,`r (skin_symptoms$severe_tab$prop[c(6,2)]*100) %>% round(1) %>% paste0(collapse="% in VIA vs. ")`%, `r skin_symptoms$severe_tab[c(2),4]`; `r pval(skin_symptoms$masto_chisq$p.value)`,  Fig. \@ref(fig:figskinsymptoms)B).

@@

Concomitant cardiovascular diseases were more prevalent in VIA than in non-VIA cases (`r age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% mutate(prop= n/sum(n)) %>% {.[c(2,5),4]} %>% pull() %>% compare(x = (age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% {.[c(2,5),3]} %>% pull()), type = "both")`) and were associated with higher risk of severe anaphylaxis when elicited by insects but were not relevant in non-VIA cases (Fig. \@ref(fig:figcofactors)). Interestingly, BST values were increased in patients with concomitant cardiovascular diseases, irrespectively of the reaction severity (Fig. S\@ref(fig:trypcardiocur)).

@@

The effect of using ACE-I (as well as beta-blockers) on the risk of severe anaphylaxis correlated with coexisting cardiovascular diseases. ACE-I use was, however, more often associated with cardiac arrests in all anaphylaxis cases (`r ace_tab %>% filter(q_114_cardiac_arrest =="yes") %>% dplyr::select(prop) %>% pull() %>% rev() %>% compare( x = ace_tab %>% filter(q_114_cardiac_arrest =="yes") %>% dplyr::select(n) %>% pull() %>% rev(), type = "both")`, `r pval(ace_tab %>% dplyr::select(-prop) %>% spread(q_114_cardiac_arrest,n) %>% {chisq.test(.[,2:3])$p.value})`) regardless of the elicitor  (Fig. \@ref(fig:figcofactors)C). 
Beta-blocker use was associated with a higher severity of anaphylaxis and with the onset of cardiovascular symptoms (cardiac arrest, chest pain), but was comparable between both VIA and non-VIA`r summary(beta_fit)[["coefficients"]][3,4] %>% pval()`). Surprisingly, arrhythmia was more frequently reported in patients with VIA and concomitant beta-blockers (Fig. \@ref(fig:figcofactors)C).

@@

`r table(rdb$grouping,rdb$q_160_ever_react)[,1:2][1,2]` (`r treact[1,2] %>% round(1)`%) of patients with insect allergy had experienced venom anaphylaxis in the past. If the reaction was elicited by other elicitors (i.e., non-VIA) --- previous reactions were more frequently seen (`r table(rdb$grouping,rdb$q_160_ever_react)[,1:2][2,2]`; `r treact[2,2]%>% round(1)`%, `r p %>% pval()`).  We observed `r sum(repeated_reaction_in_Reg)` patients with at least two fully-documented reactions. Out of these `r sum(greater_same[,2])` (`r (sum(greater_same[,2])/sum(repeated_reaction_in_Reg)*100)%>% round(1)`%) had insect elicited anaphylaxis and in `r greater_same[1,2]` of them (`r (greater_same[1,2]/sum(greater_same[,2])*100 )%>%round(1) `%), the following reaction was more severe than before. In `r greater_same[3,2]` (`r (greater_same[3,2]/sum(greater_same[,2])*100) %>% round(1)`%) cases, the reaction was similar in severity.

@@

We evaluated epinephrine use (administered by any route from patients themselves and medical professionals) in both ambulatory and emergency room settings.   
Patients who underwent VIA significantly less often received epinephrine treatment than in other anaphylaxis cases (`r ANAscore_matched %>% group_by(d_520_adren1,grouping) %>%   summarize(n = n()) %>% {.[3,3]}`; 26.9% vs. `r ANAscore_matched %>% group_by(d_520_adren1,grouping) %>%   summarize(n = n()) %>% {.[4,3]}`; 34.6%, `r test_adrenuse1$p.value %>% pval()`). After adjusting both groups for similar severity - the difference in epinephrine use was still significant irrespective of the administration route (`r ttadren1`, Fig \@ref(fig:adrenalineuse)B). 
<!-- Antihistaminic drugs, on the other hand, were given more often in cases of VIA when compared to non-VIA. -->
```{r }
pvaladrenprev <- adren_prev$data %>% 
  spread(key = d_520_adren1,value = n) %>% 
  filter(q_160_ever_react=="reacted previously") %>% 
  {chisq.test(.[,3:4])$p.value} %>% pval()

@@

A positive history of anaphylaxis influenced the therapy of a current episode as well. Epinephrine as a first-line treatment was given less often in VIA cases when compared to other cases **if patients did not report a previous history of anaphylaxis** (`r pvaladrenprev_no`), but in patients reporting previous reactions, there was no difference in epinephrine therapy (`r pvaladrenprev`, Fig. \@ref(fig:adrenalineuse)B). 
Similarly, there were no differences in the epinephrine use between VIA and non-VIA when only severe reactions were taken into consideration (`r adren_severity_sev`). However, when we restricted the analysis to moderate anaphylaxis cases --- non-VIA patients received epinephrine more frequently than VIA (`r adren_severity_mild`).  The presence of skin symptoms during these mild reactions also was associated with a lower fraction of epinephrine treated patients (Fig. \@ref(fig:rfplot)).

@@

Cofactors can promote the onset and increase the severity of anaphylaxis. 
Therefore we evaluated a variety of known cofactors regarding their impact to increase the risk of severe VIA. As expected, t

@@

The rate of concomitant cardiovascular diseases was higher in VIA than non-VIA. They are an essential cofactor increasing the risk of a severe reaction **if Hymenoptera elicited the anaphylaxis**. This association was not significant in anaphylaxis elicited by other elicitors.  Notably, cardiac arrest occurred more frequently in patients with elevated BST (> 8 ng/ml), especially in VIA. 
Nevertheless, the pathomechanism promoting cardiovascular symptoms in VIA requires further investigation.

@@

Importantly, the absence of skin symptoms was associated with more severe VIA, which was still present after excluding patients with a known diagnosis of mastocytosis. Subsequently, the correlation of BST levels with the severity of anaphylaxis lead us to identify an interaction between the absence of skin symptoms and VIA using generalized linear regression.

@@

Although there are no absolute contraindications for using epinephrine in anaphylaxis, one potential scenario where clinicians tend to be reluctant to using epinephrine is a hypersensitivity reaction presenting with high blood pressure and tachycardia, which may be present at the initial phase of VIA, due to a psychologic reflex. In theory, these less severe cases of VIA may display some form of stress-related blood pressure increase. As we lack data to confirm or discard this theory, future analysis of this question is of great clinical value.

@@

Nevertheless, because the European Anaphylaxis Registry has until now gathered over 12,000 cases of anaphylaxis - it was possible to perform a case-controlled analysis on a relatively large number of cases and investigate many aspects of VIA. It is worth underlining the important function of international registries, especially in diseases where targeted studies are not possible.

@@

# Acknowledgments

@@

```{r figsymptoms, fig.cap="Symptoms of venom-induced anaphylaxis (VIA) compared to other elicitors. A: Proportional presentation of specific reaction symptoms in VIA and non-VIA according to cardiovascular (cardio.), gastroenterologic (gastro.), and respiratory (resp.) organ systems. B: High-level overview of involved organ systems and selected cofactors in the form of a radar plot. C: difference in symptoms of VIA among patients under 22 and over 22 years of age. * denotes significant differences between groups.", fig.width=8.5, fig.height = 6.2}
fig_symptoms_3
```

@@

```{r supplement, fig.width= 7, fig.height= 5, fig.cap="Symptoms of anaphylaxis. A: The association between cardiac arrest and concomitant mastocytosis in VIA and non-VIA. B:  Hypotension frequency in two age groups of anaphylaxis. C: Crammer's V as the measure of association between groups anaphylaxis (VIA vs. non-VIA). Higher values indicate stronger association with IVA."}

@@

```{r rfplot, fig.cap="Therapy of anaphylaxis. A: Patients who presented with skin symptoms and VIA less often received epinephrine than if skin symptoms were absent during the reaction. B: Variable importance in the unsupervised classification between VIA and non-VIA using Random Forest classifier.", fig.height=3.5, fig.width=7.1}
ggarrange(
  # adren_prev+
  #   geom_segment(aes(x = 0.5,
  #                    xend = 2.5,
  #                    y = 0.33,
  #                    yend = 0.33),
  #                color="white",
  #                linetype =2),
  skin_sympt_adrenaline,
  rf_ggplot_treatment,
  nrow = 1,
  ncol = 2,
  labels=c("A","B"),
  widths = c(1,1.3)
)
```
